---
title: "Import PDF Table into R"
author: "Leila"
date: "October 10, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

## R Session Setup


```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

#### Load function libraries that will be used

*Note:* The libraries must already be installed. If they aren't, they can be installed by running:

```{r, eval = FALSE}
install.packages(c("kableExtra",
                   "knitr",
                   "tabulizer", 
                   "dplyr", 
                   "purrr",
                   "janitor"),
                 dependencies = TRUE)
```

Libraries need only be installed once, but they need to be loaded R session.

```{r libs, warning=FALSE}
# Load function libraries

library(kableExtra) # for html formatting
library(knitr) # for html formatting
library(tabulizer) # For reading the PDFS
library(dplyr, warn.conflicts = FALSE) # For intuitive data manipulation
library(purrr) # For combining output of each page into one table
library(janitor) # For cleaning up row name formatting

```


## PDF import

```{r import}
# Specify location of PDF (can be a URL or local file path)
data_url <- 'https://www.edd.ca.gov/jobs_and_training/warn/WARN-Report-for-7-1-2017-to-06-30-2018.pdf' 

# Use the extract_tables function to scan the file and pull data tables:
output <- tabulizer::extract_tables(data_url)
```

*Note from Leila:* This took 29 seconds on my desktop. This times the task:

```{r time, eval = FALSE}
system.time(output <- tabulizer::extract_tables(data_url))
```



### Inspect extraction output

Observations: 

* Output is a list (set of items), with one list item per page
* Each list item is a matrix with 8 columns

Observed tasks for later:

* A matrix can only have one data type, so all data was imported as character strings. Type conversion will need to be done
* The column names got imported as the first row of data on the first page

```{r check}

# What type of object?
class(output)

# How many list objects?
length(output)

# What is the type of object for each list item?
lapply(output, class) %>% 
  unlist() %>%
  unique()

# What is the data type in the matrices?
lapply(output,
       function(x) class(x[1,1])) %>% unlist() %>% 
  unique()

# What's the dimension of each list item?
lapply(output, dim) %>%
  do.call(rbind, .)

# Check that the last page loaded correctly 
# We can verify by looking at the file that this is correct
output[[length(output)]] %>%        # Subset to last list item
  head(n = 1) %>%                  # Subset to first row
  as.data.frame() %>% 
  kable() %>%           # print data as table
  kable_styling()       # format the table for html output


output[[length(output)]] %>% 
  tail(n = 1) %>% 
  as.data.frame() %>%
  kable() %>%
  kable_styling()

# Print the top and bottom 3 rows of the file

# Top of first list item
head(output[[1]], n = 3) %>% 
  as.data.frame() %>%
  kable() %>%
  kable_styling()

# Bottom of last list item
tail(output[[length(output)]], n = 3) %>% 
  as.data.frame() %>% 
  kable() %>%
  kable_styling()
```

## Organize output into an analyzable format

Tasks: 

1. Set column names (they were imported as the first row of data on page 1)
1. Combine all the pages into 1 table
1. Format the columns as necessary

### 1. Set column names

```{r getnames}
# Save the first row of page 1 as a character vector. 
# Subset output to the first page, first row, all columns
column_names <- output[[1]][1,] 

# View it
# It's messy with special characters, so it will have to be cleaned later.
column_names
```



```{r setnames}
# Set the column names for each page and save as new list
output_named <- purrr::map(
  # Specify list object
  .x = output,     
  # Write a function to set col names for each list item
  .f = function(x) {  
    # Get list item
    named_table <- x           
    # Set column names
    colnames(named_table) <- column_names  
    # Return named list item as output
    return(named_table)   
    }) 
```

### 2. Combine list items into 1 table

```{r combine}
# Combine the list items
output_combined <- do.call(rbind, output_named) %>%  
  # Convert to a data frame format
  as.data.frame() %>%   
  # Remove the first row (column names)
  slice(-1)                                      

# Inspect: What data types do we have in the columns?
# All factors <fct>
glimpse(output_combined)

```

### 3. Apply formatting

```{r cleanup}

# Convert dates to dates, no_of_employees to a number
output_clean <- output_combined %>% 
  # Clean up the name syntax to be a consistent format
  clean_names() %>% 
  # Set date formats
  mutate_at(.vars = c("notice_date", "effective_date", "received_date"),
            .funs = function(x) as.Date(x, format = "%m/%d/%Y")) %>% 
  # Set numeric format
  mutate_at(.vars = "no_of_employees",
            .funs = as.numeric)

# Look at the data types after type conversion
glimpse(output_clean)

```

## Write output to csv

Unless specified otherwise, R will write to the current working directory

```{r output}
# Print path of current working directory
getwd()

# Write output
write.csv(output_clean, "filename.csv")
```


