---
title: "Import PDF Table into R"
author: "Leila"
date: "October 10, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

Anyone who has worked with data knows that collecting it into an analyzable format is the first hurdle after a project is defined. Even for analysts who are the lucky recipients of curated, clean data, this is rarely a trivial task and can easily become a time-consuming and tedious distraction from the data analysis. Luckily, since it’s such a common problem, many programming angels have written helpful function packages to assist this process.

This data import series will walk through examples which use various R packages to import data freely available online. It’s intended as a beginner series, so each post will include both the minimal code snippet to complete the task at hand + an annotated version explaining what’s going on in each line.

I’ve included in the annotated segments the exploratory code used to determine how to manipulate the data into the desired format. While this intermediate exploration wouldn’t appear in production code, it’s an unavoidable step that will look different with each new project, so I’ve included it to illustrate a more complete picture of the data import process. It is not meant to represent or be confused with exploratory data analysis (the in-depth look at the data that comes after it is prepared), though the two steps often overlap.  

## Task: Import PDF Data

This example uses the R library **tabulizer** for importing a dataset stored as a table in a PDF document.  It's adapted from a [tutorial by Troy Walters](https://www.r-bloggers.com/extracting-tables-from-pdfs-in-r-using-the-tabulizer-package/) on R-Bloggers. The sample data is from the California Employment Development Department, which publishes a yearly [list of layoffs](https://www.edd.ca.gov/jobs_and_training/Layoff_Services_WARN.htm) in the state of California that fall under the WARN act. The file used is the [2017-2018 Warn Report](https://www.edd.ca.gov/jobs_and_training/warn/WARN-Report-for-7-1-2017-to-06-30-2018.pdf), containing layoffs from July 1, 2017 through June 30, 2018.

The data is relatively clean to start with: the PDF is machine readable, and the data is in a table format.  

## R Session Setup


```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(kableExtra) # For html formatting
library(knitr) # For html formatting

```

Load function libraries that will be used. Libraries need only be installed once, but they need to be loaded each new R session.

**tabulizer** is used for its `extract_tables` function to read PDFs.
**dplyr** is used for convenient, easy-to-read data wrangling
**janitor** is used to clean up table name formatting  

```{r libs, warning=FALSE}
library(tabulizer)
library(dplyr, warn.conflicts = FALSE)
library(janitor)
```

***

# Section 1: TL;DR

> Just the essential lines to import, clean, and export the data. 

Specify location of PDF (can be a URL or local file path):  

```{r specify}
data_url <- 'https://www.edd.ca.gov/jobs_and_training/warn/WARN-Report-for-7-1-2017-to-06-30-2018.pdf' 
```

Use the extract_tables() function to scan the file and pull data tables:  

```{r extract}
output <- tabulizer::extract_tables(data_url)
```

Grab the column names, which are the first row of page 1:  

```{r getnames}
column_names <- output[[1]][1,] 
```

Set the column names in the output and save as new list:  

```{r setnames}
output_named <- lapply(
  X = output,
  FUN = function(x) {
  named_table <- x           
  colnames(named_table) <- column_names  
  return(named_table)   
  }) 
```

Combine the list items:

```{r}
output_combined <- do.call(rbind, output_named) %>%  
  as.data.frame() %>%   
  slice(-1)    
```

Clean column names, convert dates to dates, convert `no_of_employees` to a number:

```{r}
output_clean <- output_combined %>% 
  clean_names() %>% 
  mutate_at(.vars = c("notice_date",
                      "effective_date",
                      "received_date"),
            .funs = function(x) as.Date(x, format = "%m/%d/%Y")) %>% 
  mutate_at(.vars = "no_of_employees",
            .funs = as.numeric)

```

Optional: write output to a csv. By default, write.csv() will write the file to the current working directory. To view the current working directory, run `getwd()` in the console.

```{r writecsv, eval = FALSE}
write.csv(output_clean, "filename.csv")
```

### All done! Here's what the top of the table looks like:

`r head(output_clean) %>% kable() %>% kable_styling()`

***

# Detailed Process Description

> This section is a more heavily-annotated version of the code above, **plus** the exploratory code used to determine each step.

The pipe `%>%` is a convenient function loaded with the `dplyr` package (originally from the `magrittr` package). It allows us to chain tasks together in a flow. When a `%>%` appears between two functions, the output from the former function is "funneled" into the latter function as its input.

## PDF import

Specify location of PDF (can be a URL or local file path):  

```{r specify2}
data_url <- 'https://www.edd.ca.gov/jobs_and_training/warn/WARN-Report-for-7-1-2017-to-06-30-2018.pdf' 
```

Use the extract_tables() function to scan the file and pull data tables:  

```{r extract2}
output <- tabulizer::extract_tables(data_url)
```

### Exploratory: Inspect extraction output

Look at the form and contents of the output in order to determine what to do with it.

What type of object is returned by `extract_tables()`?

```{r classout}
class(output)
```

How many list objects?

```{r lengthout}
length(output)
```

The list consists of one item per page of the PDF. 

What type of object are these list items?

```{r typeout}
lapply(output, class) %>% 
  unlist() %>%
  unique()
```

Each item is a matrix. 

A matrix is a 2-dimensional array of values of uniform type. What is the data type of the matrices?

```{r typemat}
lapply(output,
       function(x) class(x[1,1])) %>% 
  unlist() %>% 
  unique()
```

What's the dimension of each list item?

After `lapply` returns a list of the dimensions of each matrix, `do.call` funnels them in as the arguments to the `rbind` function, effectively binding the rows into one table.

```{r dimmat}
lapply(output, dim) %>%
  do.call(rbind, .)
```

Except for the first and last, the matrices have 48 rows x 8 columns.

Look at the data: print the top and bottom 3 rows of the file.

* The `[[ ]]` notation specifies which list item(s) to select. `length(output)` lets us access the last list item. 
* `head` and `tail` print the top and bottom (respectively) of the object (in this case, a matrix). `n` denotes how many elements (in this case, rows) to print. 
* The `kable-` lines are for presentation only -- they stylize the data printout.

```{r printends}
head(output[[1]], n = 3) %>% 
  as.data.frame() %>%
  kable() %>%
  kable_styling()

tail(output[[length(output)]], n = 3) %>% 
  as.data.frame() %>% 
  kable() %>%
  kable_styling()
```

Quality check: Look at the first and last row of the last page to Check that it loaded correctly. (Same code structure as above).

```{r checklast}
output[[length(output)]] %>%  
  head(n = 1) %>%             
  kable() %>%           
  kable_styling()       

output[[length(output)]] %>% 
  tail(n = 1) %>% 
  kable() %>%
  kable_styling()
```

Key Observations: 

* The output of `extract_tabbles()` is a list, with one list item per page.
* Each list item is a matrix with 8 columns.

Observed tasks for later:

* A matrix can only have one data type, so all data was imported as character strings. Type conversion will need to be done.
* The column names got imported as the first row of data on the first page.

## Organize output into an analyzable format

Tasks: 

1. Set column names (they were imported as the first row of data on page 1)
1. Combine all the pages into 1 table
1. Format the columns as necessary

### 1. Set column names

Save the first row of page 1 as a character vector. 

* `[[1]]` subsets the output to the data from the first page of the PDF (a matrix)
* `[1,]` subsets the matrix to the first row. The blank after the `,` means that no columns are specified, so all columns will be returned. 

```{r getnames2}
column_names <- output[[1]][1,] 
column_names
```

Set the column names for each page and save as new list called `output_named`.

* `lapply` applies the specified function `FUN` to each list item from list `X`. 
* `function(x) {...}` defines a function which sets `column_names` as the column names of the matrix `x`, then returns the named matrix as the function output.

```{r setnames2}
output_named <- lapply(
  X = output,     
  FUN = function(x) {
    named_table <- x  
    colnames(named_table) <- column_names  
    return(named_table)   
    }) 
```

### 2. Combine list items into 1 table

We've seen this syntax a few times now:
* `slice` is a `dplyr` function that selects rows, as specified by row number.
* A negative row number denoted by `-` signifies to **drop** the specified row.
* Since the first row is the column names and not data, `slice(-1)` is used to drop the first row after combining the matrices into one table.

```{r combine2}
output_combined <- do.call(rbind, output_named) %>%  
  as.data.frame() %>%   
  slice(-1)                                      
```

Inspect the combined table. `glimpse()` lists each column, the data type, and the first few rows' worth of values of the column. Note that all columns have been converted to factors (`<fct>`).

```{r inspectall}
glimpse(output_combined)
```

### 3. Apply formatting

* `clean_names()` is a `janitor` function that returns R-compatible column names. Its default settings are to remove special characters, convert spaces in names to `_` and all characters to lowercase.
* `mutate_at()` applies a specified function `.funs` to specified columns `.vars`. Here it is used to convert the date columns to dates and `no_of_employees` to a number.

```{r cleanup2}

# Convert dates to dates, no_of_employees to a number
output_clean <- output_combined %>% 
  # Clean up the name syntax to be a consistent format
  clean_names() %>% 
  # Set date formats
  mutate_at(.vars = c("notice_date", "effective_date", "received_date"),
            .funs = function(x) as.Date(x, format = "%m/%d/%Y")) %>% 
  # Set numeric format
  mutate_at(.vars = "no_of_employees",
            .funs = as.numeric)
```

Inspect the formatted table. Note the new column names and types.

```{r inspectagain}
glimpse(output_clean)
```

## Optional: Write output to csv

Unless specified otherwise, R will write to the current working directory.

`getwd()` prints the name of the current working directory, the default output destination of `write.csv()`.

```{r getwd}
getwd()
```

```{r writeout}
write.csv(output_clean, "filename.csv")
```


